---
title: "Data Preprocessing and Exploration"
output:
  pdf_document: default
---

### DALT7016 Assignment
This document will load in the NYC Vehicle Collisions dataset (379.7 MB), preprocess and explore the data, and prepare the data for visualization by generating a new file. 


###Data Preprocessing & Exploration

```{r data, message=FALSE}
# load dataset
dir = "/Users/lucadicarlo/Documents/OxfordBrookes/classes/Semester2/DALT7016-Data Visualizations/assignments/data/"
data <- read.csv(paste0(dir,'Motor_Vehicle_Collisions_NYC_Crashes.csv'))

# fill all missing values with NA's
data[data==""]= NA

print(colnames(data))

```

Only include variables and time period of interest:


```{r resize, message=FALSE}
library(dplyr)

#data[data==NA]= "NA"


# convert crash date to date object
data$CRASH.DATE <- as.Date(data$CRASH.DATE, format="%m/%d/%Y")

# convert time to integer object (0-23) for hour
data$CRASH.TIME <- as.integer(sub("\\:.*", "", data$CRASH.TIME))


# list of variables of interest
new_cols <- c("CRASH.DATE","CRASH.TIME","LATITUDE","LONGITUDE","NUMBER.OF.PERSONS.INJURED","NUMBER.OF.PERSONS.KILLED","CONTRIBUTING.FACTOR.VEHICLE.1","CONTRIBUTING.FACTOR.VEHICLE.2")

# create new dataset
data_new <- select(data,new_cols)



# only interested in complete years (2015-2019)
data_new <- data_new[data_new$CRASH.DATE > as.Date('2014-12-31',format='%Y-%m-%d'), ]
data_new <- data_new[data_new$CRASH.DATE < as.Date('2021-1-1',format='%Y-%m-%d'), ]

# add 'day of week' to the dataset
data_new$WEEKDAY <- weekdays(as.Date(data_new$CRASH.DATE))


# print new dataset
print(head(data_new),10)
```


Examine the data and search for anomalies.

```{r replace, message=FALSE}
summary(data_new)
```

Latitude and longitude exhibit extreme values for max/min, which is not expected given the coordinates for NYC should not vary more than 1 degree of latitude or longitude.

Search for outliers within continuous variables by visualizing.

```{r,echo=FALSE}
boxplot(data_new$LATITUDE,main="Latitudes", ylab='lat')
boxplot(data_new$LONGITUDE, main="Longitudes",  ylab = "lon")
plot(data_new$NUMBER.OF.PERSONS.KILLED, main="Number of People Killed",  ylab = "# Killed",cex=0.3)
plot(data_new$NUMBER.OF.PERSONS.INJURED, main="Number of People Injured",  ylab = "# Injured",cex=0.3)
```

The above plots indicate that there are significant outliers for the coordinates, which suggests errors in the data. New York City lies roughly between a latitude of 40.5-41 degrees and a longitude of -74.5 to -73.5, therefore all boxplot outliers are errors in the data or non-NYC locations.

Otherwise, both injuries and deaths related to NYC crashes indicate no abnormalities.

Removal of the outliers for latitude and longitude.

```{r outliers, message=FALSE}
# remove non-nyc locations (errors)
data_new_nyc <- data_new[(data_new$LATITUDE>=40.48) & (data_new$LATITUDE<=41),]
data_new_nyc <- data_new_nyc[(data_new_nyc$LONGITUDE>=-74.27) & (data_new_nyc$LONGITUDE<=-73.65),]

# remove rows with missing lat/lon points
data_new_nyc = data_new_nyc[!is.na(data_new_nyc$CRASH.DATE),]

diff <- length(data_new$LATITUDE) - length(data_new_nyc$LATITUDE)
print(paste0("The number of values excluded: ", diff))
```

Now view the spread of the data after removing outliers.

```{r,echo=FALSE}
boxplot(data_new_nyc$LATITUDE,main="Latitudes", ylab='lat')
boxplot(data_new_nyc$LONGITUDE, main="Longitudes",  ylab = "lon")
```
The above plots indicate more normally distributed data across a much smaller degree of latitude and longitude.

Plot the filtered data on latitude and longitude below.


```{r,echo=FALSE}
library(ggplot2)
ggplot(data_new_nyc,aes(LONGITUDE,LATITUDE)) + geom_point(size=0.01) + ggtitle("Data points of New York City")
```

The above plot confirms that the lat/lon bounds used are correct for NYC. The shape of the data clearly outlines the 5 boroughs of NYC, and can be used to create a geographic map from. 

The #Warning above the plot indicates that there are 127,258 rows of data that are missing coordinate data. Altogether that is 10.3% of the data that cannot be visualized. Mitigations for this were explored by geocoding the coordinates using street names and cross streets, although nearly every R package for this required a paid Google API to do so ($100) or capped the number of requests you could make (2,500 cap, actually need 127,258 requests).
Therefore, this data will be excluded from the analysis and visualization process as it is not complete.


Visualize categorical variables to seach for abnormalities in the data.

```{r,echo=FALSE}
library(ggplot2)
p1<-ggplot(data_new_nyc) + geom_bar(aes(CONTRIBUTING.FACTOR.VEHICLE.1))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Contributing factor to collision 1")
p2<-ggplot(data_new_nyc) + geom_bar(aes(CONTRIBUTING.FACTOR.VEHICLE.2))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Contributing factor to collision 2")
p1
p2
```
There are entries "1" and "80" which are non-descriptive factors in this data. These will be removed. In addition there are 3 categories for "cell-phone related" collisions, and two categories for "illness". These will be combined to simplify the categorical data.

Looking at the distribution of the data, there are a large number of "Unspecified" instances, which account for a majority of collisions in NYC. Comparing between CONTRIBUTING.FACTOR.VEHICLE.1 & CONTRIBUTING.FACTOR.VEHICLE.2 it is evident that factor 2 contains FAR MORE "Unspecified" data points than factor 1. Inclusion of this parameter would inflate numbers for "Unspecified" and also provide no further insight into legitimate collision-causing factors. Therefore CONTRIBUTING.FACTOR.VEHICLE.2 will not be used in the final visualization.


```{r}
# Remove CONTRIBUTING.FACTOR.VEHICLE.2
data_new_nyc <- subset(data_new_nyc,select=-CONTRIBUTING.FACTOR.VEHICLE.2)

# Combine Illnes with Illness
data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1 <- replace(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1,data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1=="Illnes","Illness")

# Combine Cell Phone factors together
data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1 <- replace(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1,data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1=="Cell Phone (hand-held)","Cell Phone")
data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1 <- replace(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1,data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1=="Cell Phone (hand-Held)","Cell Phone")
data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1 <- replace(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1,data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1=="Cell Phone (hands-free)" ,"Cell Phone")

# Combine Drug-related collisions together
data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1 <- replace(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1,data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1=="Drugs (illegal)" ,"Drugs (Illegal)")

# Remove column "1" and "80"
data_new_nyc=data_new_nyc[data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1!="80",]
data_new_nyc=data_new_nyc[data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1!="1",]

# Remove NA values
data_new_nyc <- subset(data_new_nyc,(!is.na(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1)))
```

NA values for the "cause of collision" were ommitted from the dataset as they do not convey useful information.

```{r}
p1<-ggplot(data_new_nyc) + geom_bar(aes(CONTRIBUTING.FACTOR.VEHICLE.1))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Contributing factor to collision 1")
p1
```

The above displays the simplified dataset, with 56 categories for "cause of collision".

```{r save, message=FALSE}
# save data as new CSV file
write.csv(data_new_nyc, paste0(dir,"Motor_Vehicle_Collisions_NYC_Crashes_processed.csv"))
```

### Data Examination

```{r}
library(scales)
library(plotly)
# plot most frequent accidents
causes <-data.frame(table(data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1))
causes$perc <- causes$Freq/sum(causes$Freq)*100
# order causes by descending order
causes_order <- causes[order(causes$perc,decreasing=TRUE),]

# sum the last 35 causes and create "other" category
other <- sum(causes_order$perc[12:55])

# create new dataframe by combining "other" with first 20 rows
causes_new <- data.frame(matrix(, nrow=12, ncol=3))
causes_new$X1[1:11]<- as.vector(causes_order$Var1[1:11])
causes_new$X3[1:11] <- causes_order$perc[1:11]
causes_new[12,1] = "Other"
causes_new[12,3] = other

pie <- plot_ly(causes_new,labels=~X1,values=~X3,type="pie",showlegend = TRUE,textinfo = "percent", title="Percentage of Total Collisions by Accident Cause")
pie
```
The most frequent accident cause in the city is "Unspecified", which suggests that 27.5% of vehicle collisions do not have a reason for the collision. NYC police department should work to improve the reporting of accident causes to get a better gauge of accident types. Followed in second place is "Driver Inattention/Distraction" with 22.8% of all collisions. The two of these account for nearly 50% of all accidents. The remaining causes all constitute the other 50% and are all less than 10% individually.

```{r}
cause_injured <- aggregate(data_new_nyc$NUMBER.OF.PERSONS.INJURED, by=list(Category=data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1), FUN=sum)

p3 <- ggplot(cause_injured, aes(Category,x)) + geom_bar(stat="identity",color="blue",fill="blue")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Total Injuries by Cause") + ylab("Percentage") + xlab("Collision Cause")

p3
```
The data indicates that "Driver Distraction", "Failure to Yield to Right of Way", and "Following Too Closely" are the top 3 causes of injuries to drivers. 

```{r}
cause_killed <- aggregate(data_new_nyc$NUMBER.OF.PERSONS.KILLED, by=list(Category=data_new_nyc$CONTRIBUTING.FACTOR.VEHICLE.1), FUN=sum)

p4 <- ggplot(cause_killed, aes(Category,x)) + geom_bar(stat="identity",color="red",fill="red")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("Total Killed by Cause") + ylab("Percentage") + xlab("Collision Cause")

p4
```

Speeding is the largest killer of drivers, followed by "Failure to Yield to Right of Way", and "Traffic Control Disregarded". Surprisingly "Illness" is also a large killer.

```{r}
library(reshape2)
library(plotly)
z <-dcast(data_new_nyc, WEEKDAY ~ CRASH.TIME)
z <- z[c(2,6,7,5,1,3,4),]
z <- as.matrix(z)

times <- c("0:00","1:00","2:00","3:00","4:00","5:00","6:00","7:00","8:00","9:00","10:00","11:00","12:00","13:00","14:00","15:00","16:00",
           "17:00","18:00","19:00","20:00","21:00","22:00","23:00")
dow <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

fig <- plot_ly(z =z,type = "heatmap",showscale = FALSE)
fig <- fig %>% layout(title=list(text ="All Accident Counts by Weekday & Time",showscale = FALSE,font=list(size=10)),xaxis = list(titlefont = list(size = 10),title = "Hour",fixedrange=TRUE,  tickmode = "array",nticks = 24,tickvals = c(0:23), ticktext = times, tickfont = list(size = 8)),yaxis = list(domain=c(0.6, 1),titlefont = list(size = 10),title = "Day of Week", tickmode = "array",nticks = 7,tickvals = c(0:6), ticktext = dow,fixedrange=TRUE, tickfont = list(size = 8)))
    
fig
```
A heatmap of the total number of collisions by weekday and hour of the day indicate that most accidents occur in the evening hours of the weekday. Usually during this time there is heavy traffic congestion due to commuting. The peak hours of accident occurrence is between the hours 17:00 and 18:00 on weekdays.

```{r}
ggplot(data_new_nyc,aes(LONGITUDE,LATITUDE)) + geom_point(size=0.01,alpha = 0.01) + ggtitle("Location Frequency of Collisions by Geo-Location")
```
The map above visualizes the frequency of collisions based on geolocation. Clearly there is a larger number of incidents in Manhattan and The Bronx in comparison to Staten Island and Queens.


